Syntax and Semantics of Cedille

### 2020-12-26 docs/semantics/paper.tex
docs: semantics: prove judgment validity (aka agreement)

@@

\section{Proof of Theorem~\ref{thm:syntactic-kind-pres}}
\label{sec:proof-syntactic-kind-pres}

@@

\begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
  We may rule out the cases where \(T\) is a variable or formed by a type
  constructor, as this would contradict the assumption that \(T \leadsto_{\beta}
  T'\) for some \(T'\) (recall that \(\leadsto_\beta\) is call-by-name reduction
  for types).

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa_1}{\Gamma\vdash T \tpsynth
      \abs{\Pi}{x}{T_1}{\kappa_1} & \Gamma\vdash t \tpcheck T_1}
  \]

@@

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
  to \([T_2/X]T_1''\). By inversion of the kinding derivation, we have
  \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
  Lemma~\ref{lem:subst-class}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
  [T_2/X]\kappa_1\).

\end{proof}
\end{document}

>>>>

From this theorem and a few other lemmas (see
Appendix~\ref{sec:proof-syntactic-kind-pres}), we can show the \emph{validity}
(or \emph{agreement}) of the judgments comprising CDLE.

\begin{theorem}[Judgment validity]
  \label{thm:judge-valid}
  If \(\vdash \Gamma\) then:
  \begin{itemize}
  \item if \(\Gamma \vdash T \tpsynth \kappa\) then \(\Gamma \vdash \kappa\)

  \item if \(\Gamma \vdash t \tpsynth T\) then \(\Gamma \vdash T \tpsynth \star\)
  \end{itemize}
\end{theorem}

@@

\section{Proof of Theorems~\ref{thm:syntactic-kind-pres} and \ref{thm:judge-valid}}
\label{sec:proof-syntactic-kind-pres}

@@

\begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
  We may rule out the cases where \(T\) is a variable or formed by a type
  constructor, as this would contradict the assumption that \(T \leadsto_{\beta}
  T'\) for some \(T'\) (recall that \(\leadsto_\beta\) is call-by-name reduction
  for types).
  We omit type annotations from substitutions when they are clear from the context.

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T_1}\kappa_1}{\Gamma\vdash T \tpsynth
      \abs{\Pi}{x}{T_1}{\kappa_1} & \Gamma\vdash t \tpcheck T_1}
  \]

@@

  In the other subcase, the subject of kinding is of the form
  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
  to \([T_2/X]T_1''\). By inversion of the kinding derivation, we have
  \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
  Lemma~\ref{lem:subst-class}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
  [T_2/X]\kappa_1\).
\end{proof}

\begin{proof}[Proof (of Theorem~\ref{thm:judge-valid})]
  By induction on the assumed derivation.
  We show a few interesting cases.

  \startcase{.2cm}
  \[
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma}
  \]
  By an easy inductive argument on the assumption \(\vdash \Gamma\) and
  weakening, we obtain \(\Gamma \vdash \kappa\).

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa}
  \]
  By assumption, the first premise, and the context formation rules, we have
  \(\vdash \Gamma,x:T\).
  By the IH, we have \(\Gamma,x:T \vdash \kappa\).
  Thus we obtain \(\Gamma \vdash \abs{\Pi}{x}{T}{\kappa}\).

  By the IH on the first premise, we have \(\Gamma \vdash
  \abs{\Pi}{X}{\kappa_2}{\kappa_1}\), and by inversion this gives us \(\Gamma
  \vdash \kappa_2\), which yields \(\vdash \Gamma,X:\kappa_2\), and
  \(\Gamma,X:\kappa_2 \vdash \kappa_1\).
  By the IH on the second premise, \(\Gamma \vdash \kappa_2'\).
  By Lemma~\ref{lem:ctxt-conv-class} using the third premise,
  \(\Gamma,X:\kappa_2' \vdash \kappa_1\).
  By Lemma~\ref{lem:subst-class}, \(\Gamma \vdash [T_2/X]\kappa_1\).

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T' \leadsto^*_\beta T'}
  \]
  By the IH, \(\Gamma \vdash T \tpsynth \star\).
  By generalizing Theorem~\ref{thm:syntactic-kind-pres} to \(\leadsto^*_\beta\),
  we have \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa' \cong
  \star\).
  This means that in fact \(\kappa' = \star\), as desired.

  \startcase{.2cm}
  \[
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
  \]
  By the IH on the first premise, \(\Gamma \vdash \abs{\Pi}{x}{T'}{T} \tpsynth
  \star\).
  By inversion of this, we obtain both \(\Gamma \vdash T' \tpsynth \star\), from
  which it follows by the \(\chi\) rule and second premise that \(\Gamma \vdash
  \chi\ T'\ \mhyph\ t \tpsynth T'\), and that \(\Gamma,x:T' \vdash T \tpsynth
  \star\), from which we obtain \(\Gamma \vdash [t/x]^{T'}\ T \tpsynth \star\)
  using Lemma~\ref{lem:subst-class}

  \startcase{.2cm}
  \[
    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [\beta\{t_2\}/x]^{\texttt{Top}}\ T}
    {
      \begin{array}{cc}
        \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
        & \quad \Gamma \vdash [\beta\{t_2\}/x]^{\texttt{Top}}\ T \tpsynth \star
        \\ \Gamma \vdash t \tpsynth T'
        & T' \cong [t_1/x]T
      \end{array}
    }    
  \]
  The desired result is given by the second assumption (reading first left to right, then top to bottom).

  \startcase{.2cm}
  \[
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
    {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T}
  \]
  Give to us by assumption (the first premise).
\end{proof}

\end{document}
